---
title: "Group Project"
author: "Group 12 - Ting-Wei Lin, Xinyang Qi, Jingwen Xiao"
date: "12/5/2019"
output: html_document
---

# Splines

### Introduction

The main question we are interested in is "Does people with more physial activity
get longer sleep?". A new study concludes that people sleep significantly better
and feel more alert during the day if they get at least 150 minutes of exercise a
week. Physical activity seems to have a positive effect on people's sleeping
condition, so we want to see it from real data if our data shows the same results.

Moreover, there are many factors that can affect our sleeping condition. Like the
amount of caffeine or sugar taken in a day, age and race etc. will also potentially
affect on sleeping condition. People usually think taking more caffine will have
less sleep and people with larger age will have less sleep.

Our analysis will show whether the factors we considered are significant to
sleeping time and if the relationships between each predictors and response
variable are the same as our hypotesis.


### Datasets and variables:

NHANES 2005-2006

Questionaire - 

Sleep Disorders - SLD010H (How much sleep do you get (hr.)?), 
Physical Activity - PAD080 ( How long per day(min.)), 
PAQ520 (Compare activity w/ other same age)
Physical Activity (individual ) - PADTIMES (# of times did activity in past 30 days), 
PADDURAT (Avg duration of activity(min.))
 
Dietary - 

DR1IKCAL(Energy (kcal)), DR1ISUGR (Total sugars (gm)), DR1ICAFF (Caffeine (mg))
DR2IKCAL(Energy (kcal)), DR2ISUGR (Total sugars (gm)), DR2ICAFF (Caffeine (mg))

Demographic - 

seqn, RIDGENDR (gender), RIDAGEYR (age), RIDRETH1(race/ethnicity), INDFMINC(annual family income), RIDEXMON

### Methods 

Jingwen Xiao: R (dplyr for data processing and ggplot2 on visualization)

Tingwei Lin: R (data.table for data processing and base on visualization)

Xinyang Qi: Python

We are mainly using Linear regression and Splines to fit our data.

# {.tabset .tabset-fade .tabset-pills}

## R (data.table)
```{r, echo = TRUE, message = FALSE, results = 'hide'}
## Group Project Draft
## Author: Ting Wei Lin, tingwei@umich.edu
## Updated: Dec. 5 2019

# libraries: -------------------------------------------------------------------
library(Hmisc)
library(data.table)
library(splines)

# data: ------------------------------------------------------------------------
path = "/Users/Sabrina/Documents/2019UMICH/STATS506/group project/data/"

sleep = as.data.table(sasxport.get(paste0(path, "SLQ_D.XPT")))

physical = as.data.table(sasxport.get(paste0(path, "PAQ_D.XPT")))

physical_indv = as.data.table(sasxport.get(paste0(path, "PAQIAF_D.XPT")))

demo = as.data.table(sasxport.get(paste0(path, "DEMO_D.XPT")))

dietary1 = as.data.table(sasxport.get(paste0(path, "DR1TOT_D.XPT")))

dietary2 = as.data.table(sasxport.get(paste0(path, "DR2TOT_D.XPT")))

# 80: --------------------------------------------------------------------------

# keep the variables we need
sleep = sleep[, .(seqn, sld010h)]
physical = physical[, .(seqn, pad080, paq520)]
physical_indv = physical_indv[, .(seqn, padtimes, paddurat)]
demo = demo[, .(seqn, riagendr, ridageyr, ridreth1, indfminc, ridexmon)]
dietary1 = dietary1[, .(seqn, dr1tkcal, dr1tsugr, dr1tcaff, day = 1)]
dietary2 = dietary2[, .(seqn, dr2tkcal, dr2tsugr, dr2tcaff, day = 2)]

# rename colnames
names(dietary1) = c("seqn", "drtkcal", "drtsugr", "drtcaff", "day")
names(dietary2) = c("seqn", "drtkcal", "drtsugr", "drtcaff", "day")

# merge day 1 data
data1 = merge(sleep, physical, all = TRUE)
data1 = merge(data1, physical_indv, all = TRUE)
data1 = merge(data1, demo, all = TRUE)
data1 = merge(data1, dietary1, all = TRUE)

# merge day 2 data
data2 = merge(sleep, physical, all = TRUE)
data2 = merge(data2, physical_indv, all = TRUE)
data2 = merge(data2, demo, all = TRUE)
data2 = merge(data2, dietary2, all = TRUE)

# row bind day 1 and day 2 data
data = rbind(data1, data2)

# omit the rows that have missing values
data_naomit = data[complete.cases(data[])]

# take mean for repeated seqn
avg_value = data_naomit[, lapply(.SD, mean), by = .(seqn), 
                    .SDcols = c("sld010h", "pad080", "paq520", "padtimes",
                        "paddurat", "ridageyr", "ridreth1", "indfminc",
                        "ridexmon", "drtkcal",  "drtsugr",  "drtcaff")]


```

```{r, echo = TRUE, include = TRUE}
# fit linear model 
model = lm(sld010h ~ pad080 + padtimes + ridageyr + as.factor(ridreth1) + 
             indfminc + as.factor(ridexmon) + drtkcal + drtsugr + drtcaff, 
           data = avg_value)
summary(model)

# do spline for "padtimes" variable
spline_model = lm(sld010h ~ pad080 + bs(padtimes, df = 3) + ridageyr + 
                    as.factor(ridreth1) + indfminc + as.factor(ridexmon) + 
                    drtkcal + drtsugr + drtcaff, 
                  data = avg_value)
summary(spline_model)

# plot relationship between Activity times and sleeping hours
plot(avg_value$sld010h, avg_value$pad080, col="black"
     ,xlab="Sleep (hr.)", ylab="Activity times (min.)",
     main = "Relationship between Activity times and sleeping hours")

# plot splines results
plot(avg_value$sld010h, avg_value$padtimes, col="grey",xlab="Sleep (hr.)"
     , ylab="Padtimes (# of times did activity in past 30 days)"
     , main = "Splines Results")
fit1 = smooth.spline(avg_value$sld010h, avg_value$padtimes, df=3) 
lines(fit1,col="red",lwd=2)

```

### Result

"age", "caffine" "pad080" and "ridreth1" are significant to sleeping hours. After doing Splines (df = 3) "padtimes" is partially significant to sleeping hours, but still doesn't show much of a relationship between "padtimes" and "sld010h".

## Python
```{r setup, include = FALSE}
library(reticulate)
use_python("/usr/local/bin/python")
```

```{python, eval = FALSE}

# import package and read data
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
import xport
from patsy import dmatrices
from patsy import dmatrix
from statsmodels.stats.outliers_influence import variance_inflation_factor

data1=pd.read_csv('ALQ_D.csv')
data2=pd.read_csv('PAQ_D.csv')
data3=pd.read_csv('PAQIAF_D.csv')
data4=pd.read_csv('SLQ_D.csv')
data5=pd.read_csv('DR1TOT_D.csv')
data6=pd.read_csv('DEMO_D.csv')

data2=data2[['SEQN','PAQ520']]
data4=data4[['SEQN','SLD010H']]
data3=data3[['SEQN','PADTIMES','PADDURAT']]
data5=data5[['SEQN','DR1TKCAL','DR1TSUGR','DR1TCAFF']]
data6=data6[['SEQN','RIAGENDR','RIDAGEYR']]

data=pd.merge(data2,data3,on='SEQN',how='inner')
data=pd.merge(data,data4,on='SEQN',how='inner')
data=pd.merge(data,data5,on='SEQN',how='inner')
data=pd.merge(data,data6,on='SEQN',how='inner')

data=data[data.PAQ520 != 9]
data=data[data.PAQ520 != 7]
data=data[data.SLD010H != 77]
data=data[data.SLD010H != 99]
data=data.dropna()
data=data.groupby('SEQN').mean()
data.columns=['act_level','act_times','act_durat','sleep_time','energy','sugars','caffeine','gender','age']

data['gender']=data['gender'].apply(str)
data['act_level']=data['act_level'].apply(str)
data1 = data._get_numeric_data()

# In order to improve the model, we select variables based on their vif and Covariance matrix at first. 
# Then, using backward elimination to select variables.
# get y and X dataframes based on this regression:
y, X = dmatrices('sleep_time ~ energy + sugars + caffeine + age + act_times + act_durat -1', data1, return_type='dataframe')

# Calculate VIF
vif=pd.DataFrame()
vif["VIF Factor"] = [variance_inflation_factor(X.values,i) for i in range(X.shape[1])]
vif["features"]=X.columns

print(vif)

# Calculate covariance matrix
data1.drop('sleep_time',axis=1).corr()

# Create an OLS regression model based on the latest data.
mod = smf.ols(formula='sleep_time ~ act_times + act_durat + caffeine + age + gender', data=data)
res = mod.fit()
print(res.summary())

# From the summary, we can find that the predictor of act_durat is not significant. However, maybe it doesn't have linear relationship with sleep time. 
# So we decide to use spline regression to try to improve its fit.
transformed_x1 = dmatrix("bs(data.act_durat, df=3, degree = 3, include_intercept=False)",
                        {"data.act_durat": data.act_durat}, return_type='dataframe')

mod2 = sm.GLM(data.sleep_time, transformed_x1).fit()
print(mod2.summary())

```

## R (dplyr)

```{r, echo = TRUE, include = TRUE, eval = FALSE}

# Group Project 

# Data read and process.
library(SASxport)
library(dplyr)
library(mgcv)
alc = read.xport("C:/Users/Surface-pc/Downloads/ALQ_D.xpt")
phy = read.xport("C:/Users/Surface-pc/Downloads/PAQ_D.xpt")
phy_id = read.xport("C:/Users/Surface-pc/Downloads/PAQIAF_D.xpt")
slp = read.xport("C:/Users/Surface-pc/Downloads/SLQ_D.xpt")
demo = read.xport("C:/Users/Surface-pc/Downloads/DEMO_D.xpt")
total_d1 = read.xport("C:/Users/Surface-pc/Downloads/DR1TOT_D.xpt")
total_d2 = read.xport("C:/Users/Surface-pc/Downloads/DR2TOT_D.xpt")


alc = alc %>% transmute(seqn = SEQN, alcohol = ALQ130)
phy = phy %>% transmute(seqn = SEQN, pa_comp = PAQ520)
phy_id = phy_id %>% transmute(seqn = SEQN, pa_time = PADTIMES, pa_dur = PADDURAT)
slp = slp %>% transmute(seqn = SEQN, sleep = SLD010H)
demo = demo %>% transmute(seqn = SEQN, gender = RIAGENDR, age = RIDAGEYR, 
                          race = RIDRETH1, inc = INDFMINC, winter = RIDEXMON)
total_d1 = total_d1 %>% transmute(seqn = SEQN, energy_1 = DR1TKCAL, 
                                  sugar_1 = DR1TSUGR, caffe_1 = DR1TCAFF)
total_d2 = total_d2 %>% transmute(seqn = SEQN, energy_2 = DR2TKCAL, 
                                  sugar_2 = DR2TSUGR, caffe_2 = DR2TCAFF)
data = left_join(demo, alc, by="seqn") %>% left_join(phy_id, by="seqn") %>% 
       left_join(phy, by = "seqn") %>% left_join(total_d1, "seqn") %>% 
       left_join(slp, "seqn") %>% filter(!is.na(sleep), pa_comp%in%c(1, 2, 3), 
         alcohol!=999, sleep<77, inc<12) %>% mutate(gender = ifelse(gender==1, 1, 0), 
         Mex = ifelse(race == 1, 1, 0), Hisp = ifelse(race == 2, 1, 0), 
         NHwhite = ifelse(race == 3, 1, 0), NHblack = ifelse(race == 4, 1, 0), 
         pa_high = ifelse(pa_comp == 1, 1, 0), pa_low = ifelse(pa_comp == 2, 1, 0),
         winter = ifelse(winter == 1, 1, 0), pad = pa_time*pa_dur) %>% 
       select(-c("race","pa_comp", "pa_time", "pa_dur"))

# A version without encoding factor variables.
data1 = left_join(demo, alc, by="seqn") %>% left_join(phy_id, by="seqn") %>% 
  left_join(phy, by = "seqn") %>% left_join(total_d1, "seqn") %>% 
  left_join(slp, "seqn") %>% filter(!is.na(sleep), pa_comp%in%c(1, 2, 3), 
            alcohol!=999, sleep<77, inc<12) %>% mutate(gender = ifelse(gender==1, 1, 0), 
            winter = ifelse(winter == 1, 1, 0), pad = pa_time*pa_dur) %>% 
  select(-c("pa_time", "pa_dur"))

# Quick check: how many missing values.
colSums(apply(data, 2, is.na))

dat = data%>%filter(rowSums(apply(data,2,is.na))==0) %>% group_by(seqn) %>% 
      mutate(pad = mean(pad)) %>% distinct()
dat1 = data1%>%filter(rowSums(apply(data,2,is.na))==0) %>% group_by(seqn) %>% 
  mutate(pad = mean(pad)) %>% distinct()

# Simple OLS. (Rsq~0.06)
result0=lm(sleep~.,data=dat)
summary(result0)

# GAM with some variables discarded. (Rsq~0.07)
result=gam(sleep~as.factor(Mex)+as.factor(NHwhite)+as.factor(NHblack)+as.factor(pa_low)+
           s(age,k=3)+s(alcohol,k=3)+s(pad,k=3)+
           s(energy_1,k=3)+s(sugar_1,k=3)+s(caffe_1,k=3),data=dat)
summary(result)

result1=gam(sleep~as.factor(race)+as.factor(pa_comp)+
            s(age,k=3)+s(alcohol,k=3)+s(pad,k=3)+s(energy_1,k=3)+s(sugar_1,k=3)+
            s(caffe_1,k=3),data=dat1)
summary(result1)

# Discarded variables. Do not run.
#gender+winter+s(inc,k=3)+pa_high+Hisp+NHblack+s(seqn,bs="re")
```

### End tabset
