---
title: "STATS 506 Group Project"
author: "Ting Wei Lin, tingwei@umich.edu"
date: "12/11/2019"
output: pdf_document
---

### Description
 Using data from NHANES. Select several dataset we need and analyse whether the
 variables have influence on sleeping hours. Fitting OLS and Spline regression.

```{r, message=FALSE}
# libraries: -------------------------------------------------------------------
library(Hmisc)
library(data.table)
library(splines)
```

```{r, result = hide, message=FALSE}
# data: ------------------------------------------------------------------------
path = "/Users/Sabrina/Documents/2019UMICH/STATS506/group project/data/"

sleep = as.data.table(sasxport.get(paste0(path, "SLQ_D.XPT")))
alc =  as.data.table(sasxport.get(paste0(path, "ALQ_D.xpt")))
physical = as.data.table(sasxport.get(paste0(path, "PAQ_D.XPT")))
physical_indv = as.data.table(sasxport.get(paste0(path, "PAQIAF_D.XPT")))
demo = as.data.table(sasxport.get(paste0(path, "DEMO_D.XPT")))
dietary1 = as.data.table(sasxport.get(paste0(path, "DR1TOT_D.XPT")))
dietary2 = as.data.table(sasxport.get(paste0(path, "DR2TOT_D.XPT")))
```

Data cleaning, delete the ones that have no correct response
```{r}
# 80: --------------------------------------------------------------------------

# keep the variables we need, and filter the unreasonale data
sleep = sleep[sld010h < 77, .(seqn, sld010h)]
alc = alc[alq130 < 100,.(seqn, alq130)]
physical = physical[paq520 < 7, .(seqn, paq520)]
physical_indv = physical_indv[, .(seqn, padtimes, paddurat)]
demo = demo[indfminc < 12, .(seqn, riagendr, ridageyr, ridreth1, indfminc, ridexmon)]
dietary1 = dietary1[, .(seqn, dr1tkcal, dr1tsugr, dr1tcaff)]
# rename colnames
names(dietary1) = c("seqn", "drtkcal", "drtsugr", "drtcaff")
```

###Part1.

```{r}
# merge day 1 data
data1 = merge(sleep, physical, all = TRUE)
data1 = merge(data1, alc, all = TRUE)
data1 = merge(data1, physical_indv, all = TRUE)
data1 = merge(data1, demo, all = TRUE)
data1 = merge(data1, dietary1, all = TRUE)

# rename colnames
names(data1) = c("seqn", "sleep", "pa_comp", "alcohol", "padtimes", 
                 "paddurat", "gender", "age", "race", "inc", "winter", 
                 "energy", "sugar", "caffeine")

# omit the rows that have missing values
data1_naomit = data1[complete.cases(data1[])]
```

Timing "paddurat" and "padtimes" then divide it by 60 to get "pad".
"pad" = total activity times in past 30 days in hours.
```{r}
# compute total activity times (hr.) in past 30 days
data1_pad = data1_naomit[, .(pad = paddurat*padtimes/60)]
# cbind two dataset
data1_use = cbind(data1_naomit, data1_pad)
```

We take mean for each variables within the same seqn (id) group. 
To avoid repeated seqn appears in the dataset.
Then decode the factor variables into seperate categories.
After cleaning the data, we'll get 1769 observation for this dataset.
```{r}
# take mean for repeated seqn
data1_avg_value = data1_use[, lapply(.SD, mean), by = .(seqn), 
                    .SDcols = c("sleep", "alcohol", "pa_comp", 
                                "padtimes", "paddurat", "age", "race", 
                                "inc", "winter", "gender", "energy",
                                "sugar", "caffeine", "pad")]
# decode the factors
data_new1 = data1_avg_value[, `:=` (Mex = 1L * {race == 1},
              Hisp = 1L * {race == 2}, NHwhite = 1L * {race == 3}, 
              NHblack = 1L * {race == 4}, pa_high = 1L * {pa_comp == 1}, 
              pa_low = 1L * {pa_comp == 2}, winter = 1L * {winter == 1})]
```

Fit OLS model for all variables in the dataset.
```{r}
# fit linear model for all variables
model_lm1 = lm(sleep ~ alcohol + energy + sugar + caffeine + 
                 as.factor(gender) + age + as.factor(inc) + as.factor(winter) + 
                 as.factor(Mex) + as.factor(Hisp) + as.factor(NHwhite)+
                 as.factor(NHblack) + as.factor(pa_high) + as.factor(pa_low) + 
                 pad, data = data_new1)

summary(model_lm1)
```

We found out that the Adjusted $R^2$ is really low and it contains a lot of
variables with large p-values. Hence, we decided to do variables selection
to reduce variables in our model.

```{r}
# variable selection ( drop the variables the has large p-value)
model_lm2 = lm(sleep ~ as.factor(Mex) + as.factor(NHwhite) + as.factor(NHblack) + 
                as.factor(pa_low) + age + alcohol + pad + sugar + caffeine, 
               data = data_new1)

summary(model_lm2)
```

The Adjusted $R^2$ is still really low for our fitted model after selecting variables.
So We decided to use spline regression to improve our model fitting result.

### Part 2.

First, we look at the relarionship between only "pad" and "sleep"
```{r}
# fit model for "pad" variable
model_pad1 = lm(sleep ~ pad,  data = data_new1)
summary(model_pad1)
coef(model_pad1) # extract coefficient
```

Plot for "sleep" and "pad" to see the relationship between Activity times and sleeping hours with 1769 obs.
```{r}
# plot relationship between Activity times and sleeping hours
plot(data_new1$pad, data_new1$sleep, col="darkblue"
     ,xlab="Activity times (hr.)", ylab="Sleep (hr.)",
     main = "Relationship between Activity times and sleeping hours (1769 obs)")
# add regression line to the plot
abline(coef(model_pad1)[1], coef(model_pad1)[2], col = "red")
# add legend
legend(200, 12, legend = "Prediction Line",
      col="red", lty = 1, cex=0.8)
```

The data only has 1769 observations, so we decided to re-merge the data and select
only "sleep" and "physical_indv" data.
```{r}
# create another dataset to fit only sleep and pad
data2 = merge(sleep, physical_indv, all = TRUE)
# rename colnames
names(data2) = c("seqn", "sleep", "padtimes", "paddurat")

# omit the rows that have missing values
data2_naomit = data2[complete.cases(data2[])]
# compute total activity times (hr.) in past 30 days
data2_pad = data2_naomit[, .(pad = paddurat*padtimes/60)]
# cbind two dataset
data2_use = cbind(data2_naomit, data2_pad)

# take mean for repeated seqn
data2_avg_value = data2_use[, lapply(.SD, mean), by = .(seqn), 
                           .SDcols = c("sleep", "padtimes", "paddurat", "pad")]
```

Now we have a dataset with 3946 observations.

We look at the relarionship between only "pad" and "sleep" for larger dataset.
```{r}
# fit model for "pad" variable
model_pad2 = lm(sleep ~ pad,  data = data2_avg_value)
summary(model_pad2)
coef(model_pad2) # extract colnames
```

Plot for "sleep" and "pad" to see the relationship between Activity times and sleeping hours with 3946 obs.
```{r}
# plot relationship between Activity times and sleeping hours
plot(data2_avg_value$pad, data2_avg_value$sleep, col="darkblue"
     ,xlab="Activity times (hr.)", ylab="Sleep (hr.)",
     main = "Relationship between Activity times and sleeping hours (3946 obs)")
# add regression line to the plot
abline(coef(model_pad2)[1], coef(model_pad2)[2], col = "red")
# add legend
legend(200, 12, legend = "Prediction Line",
      col="red", lty = 1, cex=0.8)
```

Reported sleep should be an approximate value. Add a jitter so that 
more respondents get the right approximated value, and we can make "sleep" a
"continuous" variable.
```{r}
# create jitter
random = rnorm(nrow(data2_avg_value), 0, 0.1)
# add jitter to "sleep" variable
data_new2 = data2_avg_value[, .(sleep = sleep + random, pad)]
```

To compare the effect of spline on "pad" variable, we first fit linear regression 
for "sleep" and "pad", then we fit spline regression for "sleep" and "pad".

Fit linear model for "pad" and "sleep" after adding jitter to "sleep".

```{r}
# do spline for "pad" variable
model_lm3 = lm(sleep ~ pad,  data = data_new2)
summary(model_lm3)
coef(model_lm3) # extract coefficient
```

Fit Spline model for "pad" and "sleep" after adding jitter to "sleep".

```{r}
# do spline for "pad" variable
spline_model = lm(sleep ~ bs(pad, df = 3),  data = data_new2)
summary(spline_model)
```

Plot to compare regression line before and after splines.
```{r}
# plot splines results compare with the one without spline
plot(data_new2$pad, data_new2$sleep, 
     col="grey", xlab="Activity times (hr.)", ylab="Sleep (hr.) with jitter", 
     main = "Splines Regression Result")
fit1 = smooth.spline(data_new2$pad, data_new2$sleep, df=3) 
abline(coef(model_lm3)[1], coef(model_lm3)[2], col = "yellow")
lines(fit1, col="red",lwd=2)
legend(200, 10, legend=c("Spline", "Without Spline"),
       col=c("red", "yellow"), lty = 1, cex=0.8)
```

### Summary
  To sum up, people getting more physical activity doesn't lead to more sleeping
hours. However, intaking less alcohol or caffeine do increase sleeping hours.
Although some research said that getting more physical activity will lead to a
better sleeping, in our analysis, we do not find physical activity significantly
related to sleeping hours. In fact there is almost no obvious relation between
those two variables.
  After fitting activity hours (pad) to a spline regression, it did fits the model
a bit better. But the relationship between physical activity and sleeping hours is
still not obvious. So we cannot say people with more physical activity
get longer sleep in this case.

